### 2019-08-26 boek.tex
update
refactorise project

\documentclass[11pt,hidelinks]{book}

@@

% Use Unicode characters when possible
%\usepackage{xltxtra}

%\usepackage[default,scale=0.95]{opensans}
%\usepackage[utf8]{inputenc}
%\usepackage[T1]{fontenc}

@@

% scripting packages
\usepackage{ifthen}
\usepackage{blindtext}
\usepackage{todonotes}

% style packages
\usepackage{lmodern}
\usepackage{fancyhdr}
%\usepackage[]{microtype} % Sublim­i­nal re­fine­ments to­wards ty­po­graph­i­cal per­fec­tion

@@

% references packages
\usepackage[sorting=none]{biblatex}
\bibliography{library}
%
%% substitute in references
%\DeclareDatafieldSet{setall}{
%	\member[datatype=literal]
%	\member[datatype=name]
%	\member[field=journal]% journal is special since it is
%	% actually journaltitle
%}
%
%\DeclareSourcemap{
%	\maps[datatype=bibtex]{
%		\map[overwrite, foreach={setall}]{
%			% \`{\i}
%			\step[fieldsource=\regexp{$MAPLOOP},
%			match=\regexp{\x{0131}\x{0300}},
%			replace=\regexp{\x{00EC}}]
%			% \'{\i}
%			\step[fieldsource=\regexp{$MAPLOOP},
%			match=\regexp{\x{0131}\x{0301}},
%			replace=\regexp{\x{00ED}}]
%			% \^{\i}
%			\step[fieldsource=\regexp{$MAPLOOP},
%			match=\regexp{\x{0131}\x{0302}},
%			replace=\regexp{\x{00EE}}]
%			% \"{\i}
%			\step[fieldsource=\regexp{$MAPLOOP},
%			match=\regexp{\x{0131}\x{0308}},
%			replace=\regexp{\x{00EF}}]
%		}
%	}
%}

@@

\tikzset{
	thumb/.style={
		%   draw=black,
		fill=light-gray,
		text=black,
		minimum height=\thumbheight, %\thumbheight,
		text width=\thumbwidth,
		outer sep=0pt,%   outer sep=10pt,
		font=\sffamily\Huge,
	}
}
\newcommand{\oddthumb}[1]{%
	\begin{tikzpicture}[remember picture, overlay]
	\node [thumb,text centered,anchor=north east,] at ($%
	(current page.north east)-%
	(0,\thumbtopmargin+\value{letternum}*\thumbheight)%
	$) {#1};
	\end{tikzpicture}
}
\newcommand{\eventhumb}[1]{%
	\begin{tikzpicture}[remember picture, overlay]
	\node [thumb,text centered,anchor=north west,] at ($%
	(current page.north west)-%
	(0,\thumbtopmargin+\value{letternum}*\thumbheight)%
	$) {#1};
	\end{tikzpicture}
}
% create a new command to set a new lettergroup
\newcommand{\lettergroup}[1]{%
	\fancyhead[LO]{\oddthumb{#1}}%
	\fancyhead[RE]{\eventhumb{#1}}%
	\fancypagestyle{chapterstart}{%
		\renewcommand{\headrulewidth}{0pt}
		\renewcommand{\footrulewidth}{0pt}
		\fancyhf{}
		\chead{\oddthumb{#1}}% chapters start only on odd pages
		\fancyfoot[RO,LE]{\textsf{Page \thepage}} %\fancyfoot[RO,LE]{\textsf{Page \thepage}}
	}
	\thispagestyle{chapterstart}
	\stepcounter{letternum}%
}

@@

%\newcommand{\picsubf}[2]{
%	\begin{subfigure}[b]{#2}
%		\fbox{\includegraphics[width=\linewidth]{#1}}
%\end{subfigure}}
%
%\newcommand{\picsubfc}[4]{
%	\begin{subfigure}[b]{#4}
%		\fbox{\includegraphics[width=\linewidth]{#1}}
%		\caption{#2}
%		\label{#3}
%\end{subfigure}}
%
%\newcommand{\picsub}[2]{
%	\begin{subfigure}[b]{#2}
%		\includegraphics[width=\linewidth]{#1}
%\end{subfigure}}
%
%\newcommand{\picsubc}[4]{
%	\begin{subfigure}[b]{#4}
%		\includegraphics[width=\linewidth]{#1}
%		\caption{#2}
%		\label{#3}
%\end{subfigure}}

@@

\definecolor{light-gray}{gray}{0.90} %not always visible on a dell screen!!!

@@

@@

\usepackage{fontspec}

\setmainfont [Path = fonts/,
UprightFont = *-300,
ItalicFont = *-300-Italic,
BoldFont = *-700,
BoldItalicFont = *-700-Italic
]{MuseoSans}

@@

\include{ch1_introduction}

% landmark mds
% princurve
% GillespieSSA
% dyngen

\include{ch2_dynbenchmark}

\include{ch3_scorpius}

% include gng ?

\include{ch4_dyno}

\include{ch5_brood}

\include{ch6_incgraph}

\include{ch7_discussion}

>>>>

\documentclass[10pt]{book}

@@

@@

% scripting packages
\usepackage{blindtext}

% style packages
\usepackage{lmodern}
\usepackage{fancyhdr}
\usepackage{microtype} % Sublim­i­nal re­fine­ments to­wards ty­po­graph­i­cal per­fec­tion

@@

% references packages
\usepackage[sorting=none]{biblatex}
\bibliography{library}
\usepackage{bibentry}
\nobibliography*

@@

\definecolor{light-gray}{gray}{0.90}
\definecolor{med-gray}{gray}{0.60}
\tikzset{
	thumb/.style={
		%   draw=black,
		fill=light-gray,
		text=black,
		minimum height=\thumbheight, %\thumbheight,
		text width=\thumbwidth,
		outer sep=0pt,%   outer sep=10pt,
		font=\sffamily\Huge,
	}
}
\newcommand{\oddthumb}[1]{%
	\begin{tikzpicture}[remember picture, overlay]
	\node [thumb,text centered,anchor=north east,] at ($%
	(current page.north east)-%
	(0,\thumbtopmargin+\value{letternum}*\thumbheight)%
	$) {#1};
	\end{tikzpicture}
}
\newcommand{\eventhumb}[1]{%
	\begin{tikzpicture}[remember picture, overlay]
	\node [thumb,text centered,anchor=north west,] at ($%
	(current page.north west)-%
	(0,\thumbtopmargin+\value{letternum}*\thumbheight)%
	$) {#1};
	\end{tikzpicture}
}
% create a new command to set a new lettergroup
\newcommand{\lettergroup}[1]{%
	\fancyhead[LO]{\oddthumb{#1}}%
	\fancyhead[RE]{\eventhumb{#1}}%
	\fancypagestyle{chapterstart}{%
		\renewcommand{\headrulewidth}{0pt}
		\renewcommand{\footrulewidth}{0pt}
		\fancyhf{}
		\chead{\oddthumb{#1}}% chapters start only on odd pages
		\fancyfoot[RO,LE]{\textsf{Page \thepage}} %\fancyfoot[RO,LE]{\textsf{Page \thepage}}
	}
	\thispagestyle{chapterstart}
	\stepcounter{letternum}%
}

@@

@@

% set font after chapter sonny style
\usepackage{fontspec}
\setmainfont [Path = fonts/,
UprightFont = *-300,
ItalicFont = *-300-Italic,
BoldFont = *-700,
BoldItalicFont = *-700-Italic
]{MuseoSans}

@@

\hypersetup{
	colorlinks = true,
%	linkcolor = {blue},
  urlcolor = {blue},
	citecolor = {med-gray}
}

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%     CH1: INTRODUCTION     %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{Introduction} 
\chaptermark{Introduction} % short title
\label{chap:introduction}
\graphicspath{{./ch1_introduction/}}
\lettergroup{\thechapter}
\begin{quote}
	\textbf{Abstract:} 
	%Recent developments in single-cell transcriptomics have opened new opportunities for studying dynamic processes in immunology in a high-throughput and unbiased manner. Starting from a mixture of cells in different stages of a developmental process, unsupervised trajectory inference algorithms aim to automatically reconstruct the underlying developmental path that cells are following. In this review, we break down the strategies used by this novel class of methods, and organize their components into a common framework, highlighting several practical advantages and disadvantages of the individual methods. We also give an overview of new insights these methods have already provided regarding the wiring and gene regulation of cell differentiation. As the trajectory inference field is still in its infancy, we propose several future developments which will ultimately lead to a global and data-driven way of studying immune cell differentiation.
\end{quote}
\vfill
Partially adapted from:\\
\textbf{Cannoodt, R.}$^*$, Saelens, W.$^*$, and Saeys, Y. Computational methods for trajectory inference from single-cell transcriptomics. \textit{European Journal of Immunology} 46, 11 (2016), 2496--2506. \doi{10.1002/eji.201646347}.\\
Saelens, W.$^*$, \textbf{Cannoodt, R.}$^*$, Todorov, H., and Saeys, Y. A comparison of single-cell trajectory inference methods. \textit{Nature Biotechnology} 37, 5 (2019), 547--554. \doi{10.1038/s41587-019-0071-9}.\\
Todorov, H., \textbf{Cannoodt, R.}, Saelens, W., and Saeys, Y. Network Inference from Single-Cell Transcriptomic Data. \textit{Gene Regulatory Networks} (2019), 235--249. \doi{10.1007/978-1-4939-8882-2_10}. \\
{\footnotesize $^*$ Equal contribution}
\newpage
\include{ch1_introduction/content}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%        CH2: DYNGEN        %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{dyngen: simulating single cells} 
\chaptermark{dyngen: simulating single cells}
\label{chap:dyngen}
\graphicspath{{./ch2_dyngen/}}
\include{ch2_dyngen/content}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%     CH3: DYNBENCHMARK     %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{dynbenchmark: A comparison of single-cell trajectory inference methods} 
\chaptermark{dynbenchmark: A comparison of single-cell trajectory inference methods.} % short title
\label{chap:dynbenchmark}
\graphicspath{{./ch3_dynbenchmark/}}
\begin{quote}
	\textbf{Abstract:} Trajectory inference approaches analyze genome-wide omics data from thousands of single cells and computationally infer the order of these cells along developmental trajectories. Although more than 70 trajectory inference tools have already been developed, it is challenging to compare their performance because the input they require and output models they produce vary substantially. Here, we benchmark 45 of these methods on 110 real and 229 synthetic datasets for cellular ordering, topology, scalability and usability. Our results highlight the complementarity of existing tools, and that the choice of method should depend mostly on the dataset dimensions and trajectory topology. Based on these results, we develop a set of guidelines to help users select the best method for their dataset. Our freely available data and evaluation pipeline (\href{https://benchmark.dynverse.org}{benchmark.dynverse.org}) will aid in the development of improved tools designed to analyze increasingly large and complex single-cell datasets.
\end{quote}
\vfill
Adapted from:\\
Saelens, W.$^*$, \textbf{Cannoodt, R.}$^*$, Todorov, H., and Saeys, Y. A comparison of single-cell trajectory inference methods. \textit{Nature Biotechnology} 37, 5 (2019), 547--554. \doi{10.1038/s41587-019-0071-9}.\\
{\footnotesize $^*$ Equal contribution}
\newpage
\include{ch3_dynbenchmark/content}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%         CH4: DYNO         %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\include{ch4_dyno}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%       CH5: SCORPIUS       %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\include{ch5_scorpius}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%         CH6: BRED         %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\include{ch6_brood}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%       CH7: INCGRAPH       %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{incgraph: Optimising regulatory networks} 
\chaptermark{incgraph: Optimising regulatory networks.} % short title
\label{chap:incgraph}
\graphicspath{{./ch7_incgraph/}}
\begin{quote}
	\textbf{Abstract:} Graphlets are small network patterns that can be counted in order to characterise the structure of a network (topology).
	As part of a topology optimisation process, one could use graphlet counts to iteratively modify a network and keep track of the graphlet counts, in order to achieve certain topological properties.
	Up until now, however, graphlets were not suited as a metric for performing topology optimisation; when millions of minor changes are made to the network structure it becomes computationally intractable to recalculate all the graphlet counts for each of the edge modifications. \\ IncGraph is a method for calculating the differences in graphlet counts with respect to the network in its previous state, which is much more efficient than calculating the graphlet occurrences from scratch at every edge modification made.
	In comparison to static counting approaches, our findings show IncGraph reduces the execution time by several orders of magnitude.
	The usefulness of this approach was demonstrated by developing a graphlet-based metric to optimise gene regulatory networks.
	IncGraph is able to quickly quantify the topological impact of small changes to a network, which opens novel research opportunities to study changes in topologies in evolving or online networks, or develop graphlet-based criteria for topology optimisation. \\
	IncGraph is freely available as an open-source R package on CRAN (incgraph). The development version is also available on GitHub (\href{http://www.github.com/rcannood/incgraph}{rcannood/incgraph}).
\end{quote}
\vfill
Adapted from:\\
\textbf{Cannoodt, R.}, Ruyssinck, J., Ramon, J., De Preter, K., and Saeys, Y. IncGraph: Incremental graphlet counting for topology optimisation. \textit{PLOS ONE} 13, 4 (2018), e0195997. \doi{10.1371/journal.pone.0195997}.
\newpage
\include{ch7_incgraph/content}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%      CH8: DISCUSSION      %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\include{ch8_discussion}

### 2019-08-26 ch1_introduction/content.tex
refactorise project

\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{Introduction} 
\chaptermark{Introduction} % short title
\label{chap:introduction}
\lettergroup{\thechapter}

\begin{quote}
	\textbf{Abstract:}

	%Recent developments in single-cell transcriptomics have opened new opportunities for studying dynamic processes in immunology in a high-throughput and unbiased manner. Starting from a mixture of cells in different stages of a developmental process, unsupervised trajectory inference algorithms aim to automatically reconstruct the underlying developmental path that cells are following. In this review, we break down the strategies used by this novel class of methods, and organize their components into a common framework, highlighting several practical advantages and disadvantages of the individual methods. We also give an overview of new insights these methods have already provided regarding the wiring and gene regulation of cell differentiation. As the trajectory inference field is still in its infancy, we propose several future developments which will ultimately lead to a global and data-driven way of studying immune cell differentiation.
\end{quote}

\vfill

Partially adapted from:\\
\textbf{Cannoodt, R.}$^*$, Saelens, W.$^*$, and Saeys, Y. Computational methods for trajectory inference from single-cell transcriptomics. \textit{European Journal of Immunology} 46, 11 (2016), 2496--2506. \doi{10.1002/eji.201646347}.\\
Saelens, W.$^*$, \textbf{Cannoodt, R.}$^*$, Todorov, H., and Saeys, Y. A comparison of single-cell trajectory inference methods. \textit{Nature Biotechnology} 37, 5 (2019), 547--554. \doi{10.1038/s41587-019-0071-9}.\\
Todorov, H., \textbf{Cannoodt, R.}, Saelens, W., and Saeys, Y. Network Inference from Single-Cell Transcriptomic Data. \textit{Gene Regulatory Networks} (2019), 235--249. \doi{10.1007/978-1-4939-8882-2_10}. \\
{\footnotesize $^*$ Equal contribution}

\newpage

@@

\begin{figure}
	\centering\includegraphics[width=.5\linewidth]{fig/introduction/fsvm3} % TODO: change image
	\caption{
		\textbf{A prominent display of the double helix shape at the VIB FSVM building.} % TODO: expand description
	}
	\label{fig:fsvm}
\end{figure}

@@

\begin{figure}
	\centering\includegraphics[width=.5\linewidth]{fig/introduction/rnaworld} % TODO: change image
	\caption{
		\textbf{RNA world.} The postulated rise and fall of the RNA world during the evolution of life, from early self-replicating RNA to complex, RNA-controlled metabolism, to the invention of translation, followed by diversification of all modern branches of life. Image from Horning (2011) \cite{horning_rnaworld_2011}. % TODO: expand the description
	}
	\label{fig:rnaworld}
\end{figure}

@@

\begin{figure}
	\centering
	A. \includegraphics[width=.45\linewidth]{fig/introduction/profiling_convergence} 
	% also figure with cell numbers
	C. \includegraphics[width=.45\linewidth]{fig/introduction/multimodal} 
	% TODO: change image
	\caption{
		A. \textbf{Convergence of "Omics" Biology and Single-Cell Biology.} Technology that allows researchers to obtain genome-wide information from single cells is extending the boundaries of a field that has thus far been limited to the analyses of a select gene in eukaryotes. Image from Junker and van Oudenaarden (2014) \cite{junker_everycellspecial_2014}. 
		B. 
		C. scmultiomics \cite{moudgil_multimodalscrnaseq_2019}. 
		% TODO: rewrite the description
	}
	\label{fig:profiling_convergence}
\end{figure}

@@

\begin{figure}
	\centering\includegraphics[width=.5\linewidth]{fig/introduction/singlecell_tools}
	\centering\includegraphics[width=.3\linewidth]{fig/introduction/scrnatools}
	% TODO: change image
	\caption{\textbf{A. Single-cell omics allows for many new types of computational approaches.} Figure adapted from Wagner et al. (2016) \cite{wagner_revealingvectorscellular_2016}.
		B. Zappia et al. (2018) \cite{zappia_exploringsinglecellrnaseq_2018}
		% TODO: rewrite the description
	}
	\label{fig:comp_tools}
\end{figure}

@@

\begin{figure}
	\centering\includegraphics[width=\Largefigure]{fig/introduction/trajectory_inference_slim} % TODO: Update figure to match text!!!!
	\caption{
		Applications of single-cell trajectory inference methods. (A) Single-cell omics data appropriate for TI can be both obtained from an unsynchronised population of single cells (snapshot data) but also from synchronised cell populations. (B) UPDATE! (C) UPDATE!
	}
	\label{fig:trajectory_inference}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=\Largefigure]{fig/introduction/ni_bulk_vs_sc}
	\caption{Bulk expression data return the average expressions of genes among large numbers of cells. In order to infer regulatory networks from this type of data, multiple bulk profiles (resulting from time series or perturbation experiments) are required. On the other hand, sequencing the transcriptome at the single-cell level uncovers the high variability among cells, providing the necessary information to infer gene regulatory networks directly.} % TODO: update label
	\label{fig:bulk_vs_singlecell}
\end{figure}

>>>>

@@

\begin{figure}
	\centering\includegraphics[width=.5\linewidth]{fig/fsvm3} % TODO: change image
	\caption{
		\textbf{A prominent display of the double helix shape at the VIB FSVM building.} % TODO: expand description
	}
	\label{fig:fsvm}
\end{figure}

@@

\begin{figure}
	\centering\includegraphics[width=.5\linewidth]{fig/rnaworld} % TODO: change image
	\caption{
		\textbf{RNA world.} The postulated rise and fall of the RNA world during the evolution of life, from early self-replicating RNA to complex, RNA-controlled metabolism, to the invention of translation, followed by diversification of all modern branches of life. Image from Horning (2011) \cite{horning_rnaworld_2011}. % TODO: expand the description
	}
	\label{fig:rnaworld}
\end{figure}

@@

\begin{figure}
	\centering
	A. \includegraphics[width=.45\linewidth]{fig/profiling_convergence} 
	% also figure with cell numbers
	C. \includegraphics[width=.45\linewidth]{fig/multimodal} 
	% TODO: change image
	\caption{
		A. \textbf{Convergence of "Omics" Biology and Single-Cell Biology.} Technology that allows researchers to obtain genome-wide information from single cells is extending the boundaries of a field that has thus far been limited to the analyses of a select gene in eukaryotes. Image from Junker and van Oudenaarden (2014) \cite{junker_everycellspecial_2014}. 
		B. 
		C. scmultiomics \cite{moudgil_multimodalscrnaseq_2019}. 
		% TODO: rewrite the description
	}
	\label{fig:profiling_convergence}
\end{figure}

@@

\begin{figure}
	\centering\includegraphics[width=.5\linewidth]{fig/singlecell_tools}
	\centering\includegraphics[width=.3\linewidth]{fig/scrnatools}
	% TODO: change image
	\caption{\textbf{A. Single-cell omics allows for many new types of computational approaches.} Figure adapted from Wagner et al. (2016) \cite{wagner_revealingvectorscellular_2016}.
		B. Zappia et al. (2018) \cite{zappia_exploringsinglecellrnaseq_2018}
		% TODO: rewrite the description
	}
	\label{fig:comp_tools}
\end{figure}

@@

\begin{figure}
	\centering\includegraphics[width=\Largefigure]{fig/trajectory_inference_slim} % TODO: Update figure to match text!!!!
	\caption{
		Applications of single-cell trajectory inference methods. (A) Single-cell omics data appropriate for TI can be both obtained from an unsynchronised population of single cells (snapshot data) but also from synchronised cell populations. (B) UPDATE! (C) UPDATE!
	}
	\label{fig:trajectory_inference}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=\Largefigure]{fig/ni_bulk_vs_sc}
	\caption{Bulk expression data return the average expressions of genes among large numbers of cells. In order to infer regulatory networks from this type of data, multiple bulk profiles (resulting from time series or perturbation experiments) are required. On the other hand, sequencing the transcriptome at the single-cell level uncovers the high variability among cells, providing the necessary information to infer gene regulatory networks directly.} % TODO: update label
	\label{fig:bulk_vs_singlecell}
\end{figure}
